Experience Report: Haskell in Computational Biology

### 2012-02-09 mrfy_experience_report.tex
tweak formatting

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
Data.Memocombinators library to provide the memoization container (again, a 
three-dimensional array). We determined that the run-time performances of both 
approaches were virtually indistinguishable, and we also noted that the 
top-down version quite faithfully resembled the recurrence relations found in a 
textbook~\cite{durbin} or the HMMER literature~\cite{eddy}. We chose to keep 
the top-down version, which later proved to be wise when we had to track down a 
bug in one of our base cases. This bug was observed as a missing state in the 
alignment our program displayed, and we quickly discovered that one of our base 
cases was missing this state: it was returning an empty list as the base of the 
path, rather than a list containing a single node. In this, we were grateful 
for the resemblance between the mathematical description of the algorithm and 
the top-down dynamic programming approach in Haskell, which resulted in 
perspicuous code.

@@

We wrote a basic search function in only twenty-five lines of code; as one of 
its parameters we provide a SearchStrategy data type, which we use to define 
specific search techniques. We require a SearchStrategy to define four 
functions: mutate, accept, terminate, and initialize; we use the Haskell module 
system to allow sharing of some of these functions between different search 
strategies. For example, both \textit{random mutation hill climb} and 
\textit{simulated annealing} use the same initialize and mutate functions, but 
very different accept functions. In all, we implemented four search strategies: 
\textit{random mutation hill climb}, \textit{simulated annealing}, 
\textit{multistart simulated annealing}, and \textit{genetic algorithm}. We 
found the implementation of the search function and search strategies to be 
pleasant, and the resulting code to be simple and elegant.

In MRFy, the Viterbi function is called several times (once for each 
non-beta-strand region) for each search generation. Furthermore, in a 
population-based search such as \textit{multistart simulated annealing}, the 
scoring function (which itself calls Viterbi) is called many times per 
generation. Beyond traditional profiling, we found parallelization to be an 
obvious way to improve runtime performance. We found this to be trivial thanks 
to the Control.Parallel library. We pass the Viterbi function to a map over all 
non-beta-strand regions of the model and query sequence, and these regions are 
independent of one another. Similarly, we pass the scoring function to a map 
over the population of potential solutions, which are also independent. We 
merely substituted ``parmap rseq'' for map. MRFy now efficiently uses all six 
cores on an AMD workstation for a non-population-based simulated annealing 
search, and all forty-eight cores on a compute server for population-based 
searches.

>>>>

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container (again, a 
three-dimensional array). We determined that the run-time performances of both 
approaches were virtually indistinguishable, and we also noted that the 
top-down version quite faithfully resembled the recurrence relations found in a 
textbook~\cite{durbin} or the HMMER literature~\cite{eddy}. We chose to keep 
the top-down version, which later proved to be wise when we had to track down a 
bug in one of our base cases. This bug was observed as a missing state in the 
alignment our program displayed, and we quickly discovered that one of our base 
cases was missing this state: it was returning an empty list as the base of the 
path, rather than a list containing a single node. In this, we were grateful 
for the resemblance between the mathematical description of the algorithm and 
the top-down dynamic programming approach in Haskell, which resulted in 
perspicuous code.

@@

We wrote a basic search function in only twenty-five lines of code; as one of 
its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques. We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies. For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions. In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}. We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.

In MRFy, the Viterbi function is called several times (once for each 
non-beta-strand region) for each search generation. Furthermore, in a 
population-based search such as \textit{multistart simulated annealing}, the 
scoring function (which itself calls Viterbi) is called many times per 
generation. Beyond traditional profiling, we found parallelization to be an 
obvious way to improve runtime performance. We found this to be trivial thanks 
to the \texttt{Control.Parallel} library. We pass the Viterbi function to a map over all 
non-beta-strand regions of the model and query sequence, and these regions are 
independent of one another. Similarly, we pass the scoring function to a map 
over the population of potential solutions, which are also independent. We 
merely substituted \texttt{parmap rseq} for map. MRFy now efficiently uses all six 
cores on an AMD workstation for a non-population-based simulated annealing 
search, and all forty-eight cores on a compute server for population-based 
searches.
